---
title: "Basic Usage"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quick guide to relatable functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette demonstrates the basic applications of the `relate` and `relation` functions. For more detailed information, see the `relatable` help documentation with `?relate` in R or [online](https://cran.r-project.org/web/packages/relatable/relatable.pdf), as well as the [Relation Types and Restrictions](https://cran.r-project.org/web/packages/relatable/vignettes/relatable-restrictions.html) vignette (`vignette("relatable-restrictions")` in R).

### A simple key-value dictionary

For basic use, `relate` maps a vector of inputs `X` from their position in a vector of keys `A` to the corresponding value in vector `B`. `relation` returns a function that performs the same mapping for repeated usage. 

```{r key_value}
library(relatable)
# Use relate() for a one-off mapping of inputs from one vector to another
relate(c("March", "April", "May"), month.name, month.abb)

# Create a reusable dictionary function with relation().
chem_symbol <- relation(elements$Name, elements$Symbol)

chem_symbol(c("Iron", "Lithium", "Sodium"))

# Unexpected inputs return <NA> by default, but this can be changed
chem_symbol(c("Sodium", "Adamantium"))

# relate() and relation() have optional arguments to determine the type of
# output and default return values.
chem_symbol <- relation(elements$Name, elements$Symbol,
  default = "Unknown", named = TRUE)

chem_symbol(c("Sodium", "Adamantium"))
```

### Ensure expected inputs while manipulating larger data sets

When working with unfamiliar data it can be easy to forget to account for all possible values a variable might take, or worse, typographical entry errors. Using `allow_default = FALSE`, `relatable` functions can flag unexpected inputs to ensure these problems don't arise.

In the following example we use the [DWNOMINATE](https://voteview.com/about) data set assembled by Poole and Rosenthal et al, which estimates the ideological positions of US politicians. Suppose we want to create a new column for our data frame indicating political party by colour (red for Republicans, blue for Democrats):

```{r data_frames}
## Obtain data for senators in the 113th Congress, spanning 2013-2015.
US_senate_113 <- subset(
  foreign::read.dta("ftp://k7moa.com/junkord/SL01113D21_BSSE_12.DTA"),
  cong == 113
)

## Setting allow_default = FALSE ensures we will be notified of any funny inputs.
US_senate_113$colour <- relate(
  X = US_senate_113$party,
  A = c(100, 200),
  B = c("blue", "red"),
  allow_default = FALSE
)

## Woops! Looks like we forgot to allow for the two Independent Senators in the data set,
## coded as 328. Let's try again:
US_senate_113$colour <- relate(
  X = US_senate_113$party,
  A = c(100, 200, 328),
  B = c("blue", "red", "gray30"),
  allow_default = FALSE
)
```

No warnings, no worries!
```{r senate_plot}
with(
  US_senate_113,
  plot(
    x = dwnom1,
    y = dwnom2,
    col = colour,
    main = "Ideal Points of US Senators",
    xlab = "Economic Issues",
    ylab = "Social Issues"
  )
)
```